HFI in RISC-V

### 2025-04-30 paper.tex
add to related works

\section{Related Work}
% TODO dev double check if this correct on the donky and cheri work
Our work on HFI for RSIC-V builds on prior work that established HFI for x86-64 architectures \cite{HFI}. HFI provides a simple ISA extension to support secure, flexible, and efficient in-process isolation, addressing limitations of existing software-based techniques, including runtime overheads, scalability limits, Spectre vulnerability, and compatibility issues.\\
There are also several alternate approaches have been developed over the years for fine-grained isolation involving hardware-assisted techniques. Some systems have proposed extensions to the page table metadata to store a per-sandbox ID checked by hardware, such as Donky \cite{Donky} or IMIX \cite{IMIX}. Donky is a hardware-software co-design providing strong in-process isolation based on dynamic memory protection domains using hardware-backed protection keys. Its implementation on RISC-V uses the reserved top 10 bits of page-table entries for protection keys, supporting 1024 keys per process. Donky distinguishes itself by managing protection key policies entirely in userspace via a dedicated control and status register (DKRU). Access to the DKRU is protected by a hardware call gate, which leverages the RISC-V N extension. This architecture provides substantially stronger security guarantees than systems like Intel MPK, where the policy register is unprivileged, allowing Donky to shield against arbitrary code execution without needing binary inspection or rewriting. Donky facilitates fast domain switches entirely in userspace, with minimal hardware additions on RISC-V. Donky has been implemented for both x86 (using MPK emulation) and RISC-V architectures, while IMIX has only been implemented on x86.\\
In addition to this, there are also methods that make use of capability-based addressing, a scheme for controlling access to memory, which is seen most prominently in Cheri \cite{CHERI}. While initially developed as a hybrid capability model extending the 64-bit MIPS ISA and demonstrated on FPGA, the CHERI approach is described as not being specific to MIPS and is a prominent capability system relevant to discussions of advanced ISAs, including RISC-V (as referenced in related work comparisons). CHERI provides byte-granularity memory protection and enables hardware enforcement of memory safety and fault isolation. It utilizes a capability coprocessor and tagged memory, where capabilities are stored in dedicated registers and memory locations, protected by hardware tags and unforgeable manipulation instructions. Capabilities are unforgeable and define protection domains based on reachability. CHERI aims for a high level of source-code and binary compatibility relative to prior pure capability machines, supporting incremental deployment. However, adapting and recompiling application code is generally required to utilize its full protection features, and achieving comprehensive object-level memory safety involves extensive modifications throughout the software stack (OS, ABI, compiler). CHERI offers strong fine-grained protection within an address space and supports scalable userspace protection domains.\\
In summary, while Donky leverages protection keys and a userspace monitor with hardware protection for the policy register, and CHERI introduces capabilities for fine-grained, unforgeable memory references requiring significant software stack changes, our work on HFI for RISC-V focuses on a region-based mechanism, minimal hardware modifications, userspace-only operation, and explicit Spectre mitigation for flexible and efficient in-process isolation supporting both adapted and unmodified code.

>>>>

\section{Related Work}
Our work on HFI for RSIC-V builds on prior work that established HFI for x86-64 architectures \cite{HFI}. HFI provides a simple ISA extension to support secure, flexible, and efficient in-process isolation, addressing limitations of existing software-based techniques, including runtime overheads, scalability limits, Spectre vulnerability, and compatibility issues.\\
There are also several alternate approaches have been developed over the years for fine-grained isolation involving hardware-assisted techniques. Some systems have proposed extensions to the page table metadata to store a per-sandbox ID checked by hardware, such as Donky \cite{Donky} or IMIX \cite{IMIX}. Donky is a hardware-software co-design providing strong in-process isolation based on dynamic memory protection domains using hardware-backed protection keys. Its implementation on RISC-V uses the reserved top 10 bits of page-table entries for protection keys, supporting 1024 keys per process. Donky distinguishes itself by managing protection key policies entirely in userspace via a dedicated control and status register (DKRU). Access to the DKRU is protected by a hardware call gate, which leverages the RISC-V N extension. This architecture provides substantially stronger security guarantees than systems like Intel MPK, where the policy register is unprivileged, allowing Donky to shield against arbitrary code execution without needing binary inspection or rewriting. Donky facilitates fast domain switches entirely in userspace, with minimal hardware additions on RISC-V. Donky has been implemented for both x86 (using MPK emulation) and RISC-V architectures, while IMIX has only been implemented on x86.\\
IMIX (In-Process Memory Isolation Extension) introduces a minimal x86 ISA extension for isolating sensitive data within a process \cite{IMIX}. Unlike many isolation mechanisms tied to a specific defense (e.g., MPK for protection keys or CET for shadow stacks), IMIX provides a mitigation-agnostic hardware primitive for protecting security-critical data. IMIX enables memory pages to be marked as isolated using a new permission flag. These pages can only be accessed with a special instruction, smov, introduced by IMIX. This mechanism allows a wide range of memory-corruption defenses (such as CPI or CFI) to protect their internal metadata from arbitrary access by potentially compromised application code. IMIX was prototyped using Intel’s simulation infrastructure, with extensions to LLVM and Linux to support its new instruction semantics. The authors demonstrated that IMIX significantly reduces the attack surface for in-process defenses with low overhead and better flexibility than existing solutions like MPK, which are limited by unprivileged key registers and require complex mitigation workarounds to be secure.\\
In addition to this, there are also methods that make use of capability-based addressing, a scheme for controlling access to memory, which is seen most prominently in CHERI \cite{CHERI}. While initially developed as a hybrid capability model extending the 64-bit MIPS ISA and demonstrated on FPGA, the CHERI approach is described as not being specific to MIPS and is a prominent capability system relevant to discussions of advanced ISAs, including RISC-V (as referenced in related work comparisons). CHERI provides byte-granularity memory protection and enables hardware enforcement of memory safety and fault isolation. It utilizes a capability coprocessor and tagged memory, where capabilities are stored in dedicated registers and memory locations, protected by hardware tags and unforgeable manipulation instructions. Capabilities are unforgeable and define protection domains based on reachability. CHERI aims for a high level of source-code and binary compatibility relative to prior pure capability machines, supporting incremental deployment. However, adapting and recompiling application code is generally required to utilize its full protection features, and achieving comprehensive object-level memory safety involves extensive modifications throughout the software stack (OS, ABI, compiler). CHERI offers strong fine-grained protection within an address space and supports scalable userspace protection domains.\\
In summary, while Donky leverages protection keys and a userspace monitor with hardware protection for the policy register, and CHERI introduces capabilities for fine-grained, unforgeable memory references requiring significant software stack changes, IMIX proposes a lightweight ISA extension with minimal hardware cost to securely restrict access to sensitive memory used by diverse in-process defenses. Our work on HFI for RISC-V focuses on a region-based mechanism, minimal hardware modifications, userspace-only operation, and explicit Spectre mitigation for flexible and efficient in-process isolation supporting both adapted and unmodified code.
